---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Ben Hardin"
date: "2/14/2023"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r prep-data}
#reading data
staff <- read.csv("data/instructional-staff.csv")

#clean the years
staff <- staff %>%
  rename("1975" = X1975, "1989" = X1989, "1993" = X1993, "1995" = X1995, "1999" = X1999, "2001" = X2001, "2003" = X2003, "2005" = X2005, "2007" = X2007, "2009" = X2009, "2011" = X2011)

#reshaping data
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

### Exercise 1

Here is a new graph displaying how the proportion of different faculty types making up new hires has changed over time.

```{r making-first-plot, fig.height = 4, fig.width = 8}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type))+
  geom_line(size = 1.1)+
  geom_point(aes(shape = faculty_type), size = 1.5)+
  theme_bw()+
  labs(title = "Trends in faculty hires (1975 - 2011)",
       y = "Percentage of faculty hires",
       x = "Year",
       color = "Faculty type",
       shape = "Faculty type")
```

### Exercise 2

I do think the plot does an ok job displaying how the proportion of part-time faculty has increased dramatically compared to other instructors, but this piece of information is obscured a little bit by the amount of things happening in the plot, and that they are all about equally prominent in the graph. For that reason, I want to propose making 2 changes to the plot.

1. Combining all full-time faculty into 1 category, so that we are only comparing 3 kinds of instructor types and not cluttering the graph quite so much.
2. Making the line for part-time faculty brighter and more colorful than the other lines, so that our attention is drawn to the instructor type that we are most interested in learning about.

```{r making-transforming-data}
staff_long <-  staff_long %>%
  mutate(full_time = case_when(
    faculty_type == "Part-Time Faculty" ~ "Part-Time Faculty",
    faculty_type == "Graduate Student Employees" ~ "Graduate Student Employees",
    faculty_type %in% c("Full-Time Tenured Faculty", "Full-Time Tenure-Track Faculty", "Full-Time Non-Tenure-Track Faculty") ~ "Full-Time Faculty"))

staff_full <- aggregate(value ~ year + full_time, data = staff_long, FUN = sum)
```

```{r making-new-plot}
ggplot(data = staff_full, aes(x = year, y = value, group = full_time, color = full_time))+
  geom_line(size = 1.1)+
  geom_point(aes(shape = full_time), size = 1.75)+
  theme_bw()+
  scale_color_manual(values = c("Full-Time Faculty" = "gray50", "Graduate Student Employees" = "gray80", "Part-Time Faculty" = "orange"))+
  labs(title = "Trends in faculty hires (1975 - 2011)",
       y = "Percentage of faculty hires",
       x = "Year",
       color = "Faculty type",
       shape = "Faculty type")
```

I'm pretty pleased with the result! In particular, it highlights something pretty striking that would have been really hard to deduce from the previous graph -- that since 2007, the proportion of part-time faculty hires has actually exceeded the proportion of all full-time faculty types combined! :0

### Exercise 3

```{r doin-stuff}
fisheries <- read_csv("data/fisheries.csv")

fisheries %>%
  mutate(new_total = total/1000000) %>%
  filter(total > 100000)%>%
ggplot(aes(x = new_total))+
  geom_histogram(binwidth = 1, color = "black", fill = "blue")+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25))+
  scale_x_continuous(expand = c(0, 0), limits = c(0, 82))+
  theme_bw()
```
```{r try-something-better}
fish_long <- fisheries %>%
  pivot_longer(cols = c(-total, -country), names_to = "type") %>%
  mutate(value = as.numeric(value))

fish_long %>%
  mutate(new_country = case_when(
    country == "China" ~ "China",
    country == "Indonesia" ~ "Indonesia",
    country == "India" ~ "India",
    country == "Vietnam" ~ "Vietnam",
    country == "United States" ~ "United States",
    country == "Russia" ~ "Russia",
    !country %in% c("China", "India", "Indonesia", "Vietnam", "United States", "Russia") ~ "Other")) %>%
  ggplot(aes(x = type, y = value, fill = fct_reorder(new_country, total)))+
  geom_bar(stat = "identity")
```