---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Ben Hardin"
date: "2/14/2023"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r prep-data}
#reading data
staff <- read.csv("data/instructional-staff.csv")

#clean the years
staff <- staff %>%
  rename("1975" = X1975, "1989" = X1989, "1993" = X1993, "1995" = X1995, "1999" = X1999, "2001" = X2001, "2003" = X2003, "2005" = X2005, "2007" = X2007, "2009" = X2009, "2011" = X2011)

#reshaping data
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

### Exercise 1

Here is a new graph displaying how the proportion of different faculty types making up new hires has changed over time.

```{r making-first-plot, fig.height = 4, fig.width = 8}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type))+
  geom_line(size = 1.1)+
  geom_point(aes(shape = faculty_type), size = 1.5)+
  theme_bw()+
  labs(title = "Trends in faculty hires (1975 - 2011)",
       y = "Percentage of faculty hires",
       x = "Year",
       color = "Faculty type",
       shape = "Faculty type")
```

### Exercise 2

I do think the plot does an ok job displaying how the proportion of part-time faculty has increased dramatically compared to other instructors, but this piece of information is obscured a little bit by the amount of things happening in the plot, and that they are all about equally prominent in the graph. For that reason, I want to propose making 2 changes to the plot.

1. Combining all full-time faculty into 1 category, so that we are only comparing 3 kinds of instructor types and not cluttering the graph quite so much.
2. Making the line for part-time faculty brighter and more colorful than the other lines, so that our attention is drawn to the instructor type that we are most interested in learning about.

```{r making-transforming-data}
staff_long <-  staff_long %>%
  mutate(full_time = case_when(
    faculty_type == "Part-Time Faculty" ~ "Part-Time Faculty",
    faculty_type == "Graduate Student Employees" ~ "Graduate Student Employees",
    faculty_type %in% c("Full-Time Tenured Faculty", "Full-Time Tenure-Track Faculty", "Full-Time Non-Tenure-Track Faculty") ~ "Full-Time Faculty"))

staff_full <- aggregate(value ~ year + full_time, data = staff_long, FUN = sum)
```

```{r making-new-plot}
ggplot(data = staff_full, aes(x = year, y = value, group = full_time, color = full_time))+
  geom_line(size = 1.1)+
  geom_point(aes(shape = full_time), size = 1.75)+
  theme_bw()+
  scale_color_manual(values = c("Full-Time Faculty" = "gray50", "Graduate Student Employees" = "gray80", "Part-Time Faculty" = "orange"))+
  labs(title = "Trends in faculty hires (1975 - 2011)",
       y = "Percentage of faculty hires",
       x = "Year",
       color = "Faculty type",
       shape = "Faculty type")
```

I'm pretty pleased with the result! In particular, it highlights something pretty striking that would have been really hard to deduce from the previous graph -- that since 2007, the proportion of part-time faculty hires has actually exceeded the proportion of all full-time faculty types combined! :0

### Exercise 3

```{r doin-stuff}
fisheries <- read_csv("data/fisheries.csv")

fisheries %>%
  mutate(new_total = total/1000000) %>%
  filter(total > 100000)%>%
ggplot(aes(x = new_total))+
  geom_histogram(binwidth = 1, color = "black", fill = "blue")+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25))+
  scale_x_continuous(expand = c(0, 0), limits = c(0, 82))+
  theme_bw()
```
```{r try-something-better}
fish_long <- fisheries %>%
  pivot_longer(cols = c(-total, -country), names_to = "type") %>%
  mutate(value = as.numeric(value))

fish_long %>%
  mutate(new_country = case_when(
    country == "China" ~ "China",
    country == "Indonesia" ~ "Indonesia",
    country == "India" ~ "India",
    country == "Vietnam" ~ "Vietnam",
    country == "United States" ~ "United States",
    country == "Russia" ~ "Russia",
    !country %in% c("China", "India", "Indonesia", "Vietnam", "United States", "Russia") ~ "Other")) %>%
  ggplot(aes(x = type, y = value, fill = fct_reorder(new_country, total)))+
  geom_bar(stat = "identity")

fish_long %>%
  filter(!country %in% c("China", "India", "Indonesia", "Vietnam", "United States", "Russia"), total > 1000000) %>%
  ggplot(aes(x = type, y = value, fill = fct_reorder(country, value)))+
  geom_bar(stat = "identity")
```

```{r one-last-plot}
fish_long %>%
  filter(total > 1000000) %>%
  ggplot(aes(x = fct_reorder(country, value), fill = type, y = value))+
  geom_bar(stat = "identity")
```

### Stretch Exercises
```{r load-library}
library(mosaicData)

data(Whickham)
```
### Exercise 1

The data is most likely observational, because it is described as a survey conducted at 2 time points, and also because the IV is whether someone is a smoker. Given the health risks involved in smoking, it is unlikely that the researchers would have randomly assigned people to become smokers for 20 years, which would be extremely unethical.

### Exercise 2

There are 1314 observations in the dataset, each of which represents a woman in the town of Whickham.

### Exercise 3

There are 3 variables in the dataset, which represent:

1. Whether the participant was still alive 20 years after the initial survey (dichotomous)
2. The participants smoker status at the time of the initial survey (dichotomous)
3. The participants age at the time of the initial survey (continuous)

```{r viz-vars}
ggplot(Whickham, aes(x = outcome))+
  geom_bar()

ggplot(Whickham, aes(x = smoker))+
  geom_bar()

ggplot(Whickham, aes(x = age))+
  geom_histogram(color = "black", fill = "grey80")
```

### Exercise 4

My expectation would be that smoking would be associated with worse health outcomes. In this case, that would mean greater or earlier mortality among smokers.

### Exercise 5

Based on the visualization, people who reported being smokers at time 1 don't seem to differ much from people who reported not being smokers in terms of their likelihood of surviving to time 2. The conditional probabilities reveal that, apparently contrary to my prediction, smokers were descriptively less likely to have died by time 2 than non-smokers.

```{r viz-results}
ggplot(Whickham, aes(x = smoker, fill = outcome))+
  geom_bar(position = "fill")
```

```{r condit-prob}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(prob_dead = n/sum(n))
```

### Exercise 6

```{r new-var}
Whickham <- Whickham %>%
  mutate(age_cat = case_when(
    age <= 44 ~ "18 - 44",
    age > 44 & age <= 64 ~ "45 - 64",
    age > 64 ~ "65+"))
```

### Exercise 7

Not too surprisingly, the likelihood of mortality was much higher for people who were 65 and above at time 1, and much lower for people between 18-44. For younger people, the those who were smokers were somewhat more likely to have died by time 2 than non-smokers. For 65 and older people, if we were to only focus on comparing the "dead" columns, it would look like more non-smokers than smokers are dying. However, when we compare the size of the bars, we can see that very few smokers above the age of 65 were ever surveyed, and a higher proportion of them died compared to non-smokers over 65. The original plot was misleading about the effect, because a majority of 65+ individuals died regardless of their smoker status and there weren't very many 65+ smokers available to sample in the first place. Combining all the age categories together masked these nuances of the data. This is an example of Simpson's paradox.

```{r unraveling-age}
ggplot(Whickham, aes(x = smoker, fill = outcome))+
  facet_wrap(~ age_cat)+
  geom_bar()

Whickham %>%
  count(smoker, outcome, age_cat) %>%
  pivot_wider(names_from = age_cat, values_from = n)
```